\documentclass[12pt]{article}
\RequirePackage[l2tabu, orthodox]{nag}
\usepackage[main=english]{babel}
\usepackage[rm={lining,tabular},sf={lining,tabular},tt={lining,tabular,monowidth}]{cfr-lm}
\usepackage{amsthm,amssymb,latexsym,gensymb,mathtools,mathrsfs}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[pdftex]{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{epstopdf,enumitem,microtype,dcolumn,booktabs,hyperref,url,fancyhdr}
\usepackage{bbm}

% Plotting
\usepackage{pgfplots}
\usepackage{xinttools} % for the \xintFor***
\usepgfplotslibrary{fillbetween}
\pgfplotsset{compat=1.8}
\usepackage{tikz}

% Local custom commands. 
\include{local-defs-GPs}

\setlist{topsep=1ex,parsep=1ex,itemsep=0ex}
\setlist[1]{leftmargin=\parindent}
\setlist[enumerate,1]{label=\arabic*.,ref=\arabic*}
\setlist[enumerate,2]{label=(\alph*),ref=(\alph*)}

% Specifically for paper formatting 
\renewcommand{\baselinestretch}{1.2} % Spaces manuscript for easy reading

% Formatting definitions, propositions, etc. 
\newtheorem{definition}{Definition}
\newtheorem{prop}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{thm}{Theorem}
\newtheorem{corollary}{Corollary}

\begin{document}

\begin{center}
\Large
Measure Transport, Normalizing Flows, and Diffusion Models 
\end{center}

\begin{flushright}
Andrew Roberts
\end{flushright} 

% Background: Change of Variables
\section{Background: Change of Variables}

\subsection{Integration}
TODO: use Hubbard and Hubbard 

\subsection{Probability}


% Bibliography
\begin{thebibliography}{20}
\bibitem{Papamakarios} George Papamakarios, Eric Nalisnick, Danilo Jimenez Rezende, Shakir Mohamed, and Balaji Lakshminarayanan. 2021. Normalizing flows for probabilistic modeling and inference. J. Mach. Learn. Res. 22, 1, Article 57 (January 2021), 64 pages.
\bibitem{Kobyzev} I. Kobyzev, S. Prince and M. Brubaker, "Normalizing Flows: An Introduction and Review of Current Methods" in IEEE Transactions on Pattern Analysis \& Machine Intelligence, vol. 43, no. 11, pp. 3964-3979, 2021. doi: 10.1109/TPAMI.2020.2992934
\bibitem{Marzouk} Marzouk, Youssef \& Moselhy, Tarek \& Parno, Matthew \& Spantini, Alessio. (2016). Sampling via Measure Transport: An Introduction. 

\bibitem{Moselhy} Tarek A. El Moselhy, Youssef M. Marzouk, Bayesian inference with optimal maps, Journal of Computational Physics, Volume 231, Issue 23, 2012, Pages 7815-7850, ISSN 0021-9991, https://doi.org/10.1016/j.jcp.2012.07.022.

\bibitem{Song} Yang Song and Jascha Sohl-Dickstein and Diederik P. Kingma and Abhishek Kumar and Stefano Ermon and Ben Poole. Score-Based Generative Modeling through Stochastic Differential Equations, 2021. 

\bibitem{Simpson} Dan Simpson. 2023. “Diffusion Models; or Yet Another Way to Sample from an Arbitrary Distribution.” February 9, 2023. https://dansblog.netlify.app/posts/.

\bibitem{Baptista} Ricardo Baptista, Probabilistic modeling and Bayesian inference via triangular transport. PhD Thesis, 2022. 

\end{thebibliography}

\end{document}


